---
title: "timeaveraging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{timeaveraging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**INTRODUCTION**

Package *timeaveraging* assembles functions for analyzing age distribution of individually dated fossil specimens. The package is designed for datasets that include posterior-calibrated ages (typically Bayesian-calibrated radiocarbon or amino-acid racemization dates). Package includes example datasets are included.

The current version of the package time-averaging can be downloaded from Github.

```{r installing timeavraging package, eval = FALSE}
install.packages('devtools')
library(devtools)
devtools::install_github('mjkowalewski/timeaveraging', build_vignettes = TRUE)
```

```{r}
library(timeaveraging)
```


**FUNCTIONS**

*postdist* - generates specimen-level Monte Carlo distributions of age estimates for each dated specimen based on Bayesian posterior probability distribution. Returns a list or matrix of class 'postdist'.

*tastats* - computes sample-level statistics of age uncertainties based on Bayesian posterior probabilities of ages

**DATASETS**

*bahamas* - A dataset including radiocarbon ages for 30 mollusk shells and 30 echinoid tests from one sampling site off the coast of San Salvador, Bahamas. Data from Kowalewski et al. (2018).

*florida* - An unpublished dataset of radiocarbon ages for mollusk shells, echinoid tests, mollusk fragments, and echinoid samples from Cedar Key, Florida.



**EXPLANATIONS AND EXAMPLES**

*postdist* 

The function *postdist* generates specimen-level Monte Carlo distributions of age estimates for each dated specimen based on Bayesian posterior probability distribution. Returns a list or matrix of class 'postdist'.

NOTE: Radiocarbon calibration provides posterior probabilities of age values given the observed uncalibrated radiocarbon ages. The function *postdist* generates a Monte Carlo approximation of posterior probability distribution converting probabilities into a predefined number of age values representing age uncertainty around a given date. If age bins are not annual, the function adjust probabilities to ensure they sum up to 1.

Arguments:

*age* - A numerical vector containing age-bin values to which probabilities are assigned

*prob* - A numerical vector containing probabilities for each age value

*id* - A vector containing unique specimen ids

*size* - a single numerical value (default = 10000) defining how many age values should be sampled to approximate the posterior probability distribution.

Returns a single matrix of the class 'postdist' with rows representing individual age dates columns representing Monte Carlo ages sampled from posterior probability distributions. Output data are sorted by specimen id.

```{r}
out1 <- postdist(age=bahamas$post$Age, prob=bahamas$post$Probability, id=bahamas$post$Specimen)
dim(out1)
out1[1:5,1:10]
```

**tastats FUNCTION**

The function *tastats* returns summary of statistics describing an age distribution of age dates. Accepts object of the class 'postidst' generated by function *postdist*.

If argument 'simple' is set to TRUE (default), the function returns a vector with summary statistics for the analyzed sample. If argument 'simple' is set to FALSE, the function returns a list

The function returns the following summary statistics:

*median.age* - median age of a sample calculated from mean or median ages of posterior distributions of individual ages

*mean.age* - mean age of a sample calculated from mean or median ages of posterior distributions of individual ages.

*std.age* - standard deviation of a sample calculated from mean or median ages of posterior distributions of individual ages.

*iq.age* - interquartile range calculated from mean or median ages of posterior distributions of individual ages.

*AEV* - Age Estimation Variance calculated as average of variances of posterior Monte-Carlo distributions of individual ages included in a sample (see Ritter et al. 2023)

*TAV* - Total Assemblage Variance calculated as variance of all Monte-Carlo ages (see Ritter et al. 2023)

*ETA* - Estimated Time-Averaging (2 standard deviations) corrected for dating uncertainty calculated as 2 * square root of TAV-AEV (see Ritter et al. 2023)

*g1* - skewness, the third moment about the mean calculated from mean or median ages of posterior distributions.

*l3* - L-skeweness calculated using function *samlmu* from the package *lmom* (Hosking 2024).

Arguments:

*x* - A list of class 'postdist' returned by *postdist* function

*use.median* - A logical argument (default=TRUE) to determine whether to use median (TRUE) or mean (FALSE) to estimate central tendency of the posterior distribution of age probabilities

*simple* - A logical argument (default=TRUE) to determine if additional information (simple=FALSE) should be returned by the function.

Values:

NOTE: The function *tastats* return a dataframe with sample-level statistics (if simple=T) or a list with five items (if simple=F).

*my.stats* - a dataframe (if simple = T) or the first item in the list of items (if simple = F). This is a single set of numerical values providing summary statistics for a given sample.

*ages* - age estimates (mean or median ages) for individual observations (returned if simple = F)

*n* - the number of age dates in a sample (returned if simple = F)

*num.replic.ages* - the number of Monet Carlo replicate samples from the posterior probability distibution of each age date (returned if simple = F)

*age.estimate* - returns the information if median or mean of the posterior probability distribution was used to derive an estimated specimen age (returned if simple = F)

```{r}
out1 <- postdist(age=florida$post$cal.BP, prob=florida$post$probability, id=florida$post$UniqueID, size=10000)
# NOTE: Default version returns a one-column matrix
out2 <- tastats(out1)
out2
# NOTE: when the argument simple is set to FALSE, the function returns a list with five items
out3 <- tastats(out1, simple=F)
str(out3)
out3$my.stats
hist(out3$ages, main='', xlab='calibrated age [BP]', ylab='number of specimens')
```

*Comments*
It is advisable to check if Monte Calo estimates agree with the median ages (if provied in the specimen dataset)

```{r}
out1 <- postdist(age=bahamas$post$Age, prob=bahamas$post$Probability,
                 id=bahamas$post$Specimen)
out2 <- postdist(age=florida$post$cal.BP, prob=florida$post$probability,
                 id=florida$post$UniqueID)
bout <- data.frame(id=bahamas$spec$sample[order(bahamas$spec$sample)],
           age=bahamas$spec$medianAge[order(bahamas$spec$sample)], agemc=apply(out1,1,median))
fout <- data.frame(id=florida$spec$UniqueID[order(florida$spec$UniqueID)],
           age=florida$spec$med_cal_age[order(florida$spec$UniqueID)], agemc=apply(out2,1,median))
plot(fout$age, fout$agemc, ylab='Monte Carlo estimate', xlab='reported calibrated median age')
 abline(a=0, b=1)
plot(bout$age, bout$agemc, ylab='Monte Carlo estimate', xlab='reported calibrated median age')
 abline(a=0, b=1)
```


**DATASETS**

**bahamas DATASET**

A dataset including radiocarbon ages for 30 mollusk shells and 30 echinoid tests from one sampling site off the coast of San Salvador, Bahamas. Data from Kowalewski et al. (2018).

*$spec* - dataframe with a specimen-level dataset (rows = specimens) with variables (columns) describing each specimen. 

*$post* - posterior probability distributions for each specimen. Rows are individual age values (0.5-year resolution bins) and variables include: Specimen - specimen id , Age - age-bin value, Probability - age-bin probability, Taxon - taxonomic assignment of a given specimens.

```{r}
str(bahamas)
```

```{r}
bahamas$post[1:5,]
```

**florida DATASET**

An unpublished dataset of radiocarbon ages for mollusk shells, echinoid tests, mollusk fragments, and echinoid samples from Cedar Key, Florida. This is a list that includes two dataframes.

*$spec* - dataframe with a specimen-level dataset (rows = specimens) with variables (columns) describing each specimen. 

*$post* - posterior probability distributions for each specimen. Rows are individual age values (0.5-year resolution bins) and variables include: index - index number, op - type of age estimate, UniqueID - specimen id , type - estimate type, cal.BP - age-bin value, probability - age-bin probability.

```{r}
str(florida)
```

```{r}
florida$post[1:5,]
```

**REFERENCES**

Hosking, J.R.M., (2024), L-Moments. R package, version 3.2. (https://CRAN.R-project.org/package=lmom)

Kowalewski, M., Casebolt*, S., Hua, Q., Whitacre, K.E., Kaufman, D., Kosnik, M.A., (2018), One fossil record, multiple time-resolutions: Comparative time-averaging of mollusks and echinoids on a carbonate platform. Geology 46: 51-54. (https://doi.org/10.1130/G39789.1)

Ritter, M, Erthal F, Kosnik, MA, Kowalewski, M, Coimbra, CJ, Caron, F, Kaufman, D.S., (2023), Onshore-offshore trends in the temporal resolution of molluscan death assemblages: how age-frequency distributions reveal Quaternary sea-level history. Palaios, 38: 148-157. (http://dx.doi.org/10.2110/palo.2021.041)
