---
title: "timeaveraging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{timeaveraging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Package time-averaging assembles functions for analyzing age distribution of individually dated fossil specimens. The package is designed for datasets that include posteriorly calibrated ages (typically calibrated radiocarbon or amino-acid racemization dates). Example datasets are included.

The current version of the package time-averaging can be downloaded from Github:
devtools::install_github('mjkowalewski/timeaveraging', build_vignettes = TRUE)

```{r setup}
library(timeaveraging)
```

The following functions are included:
*agemoments* - a function that provides basic descriptors of age distributions and can be used with other functions

The following example datasets are provided:
*bahamas* - a list with two dataframes (spec - specimen level data, and post - posterior probability distributions for each specimen) for 30 specimens of echinoids and 30 specimens of mollusks.

