---
title: "timeaveraging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{timeaveraging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Package time-averaging assembles functions for analyzing age distribution of individually dated fossil specimens. The package is designed for datasets that include posteriorly calibrated ages (typically calibrated radiocarbon or amino-acid racemization dates). Example datasets are included.

The current version of the package time-averaging can be downloaded from Github.

```{r installing timeavraging package, eval = FALSE}
install.packages('devtools')
library(devtools)
devtools::install_github('mjkowalewski/timeaveraging', build_vignettes = TRUE)
library(PaleoFidelity)
```

The following functions and datasets are included:

*postdist* - a function that creates Monte Carlo distributions from posterior probabilities and provides basic descriptors of specimen-level posterior distributions and sample-level statistics of time-averaging. Retruns a list of class 'postdist'

*bahamas* - a list with two dataframes

$spec - specimen level data

$post - posterior probability distributions for each specimen (30 echinoid specimens and 30 mollusk specimens).

